### 2013-05-31 doc/nips2013/abstract.tex
made a pass up through section 4 (measurements)

\begin{abstract}
Latent-variable models.
\end{abstract}

>>>>

\begin{abstract}
Parameter estimation in log-linear latent-variable models is typically
performed using gradient-based optimization on the log-likelihood, which is prone to local optima.
In this paper,
%we show how to leverage recent advances in the method of moments
%to alleviate these problems.
we leverage recent advances in the method of moments to estimate latent
moments of the target distribution,
which we incorporate as constraints on the original optimization
problem.
In favorable cases, these constraints provide full information and render the
original optimization problem
convex; in other cases, the constraints offer only partial information,
but still appear to provide significant aid to the original optimization
problem.
Factorial models do require special attention;
to handle these models,
we introduce a new technique called unshuffling factorization to induce
linear constraints on the parameters.
Empirically, our approach provides improvements on several models.
\end{abstract}

### 2013-05-31 doc/nips2013/introduction.tex
made a pass up through section 4 (measurements)

% Method of moments
Recently, unsupervised learning techniques based on the method of moments and
spectral decomposition have offered a refreshing and promising perspective on
this learning problem \citep{hsu09spectral,anandkumar11tree,anandkumar12moments,anandkumar12lda,hsu12identifiability,balle11transducer,balle12automata}.
These methods exploit the linear algebraic properties of the model and
factorize the moments into parameters, providing strong theoretical guarantees.
However, these methods are not as easy to apply as EM, and have yet to be
developed for log-linear models.

@@

\Fig{figures/schema}{0.3}{schema}{Schema.}

This paper contributes three ideas:
First, in certain cases, these latent moments are exactly the sufficient statistics of
the log-linear model, in which we case use convex optimization directly to obtain the parameters,
resulting in a consistent parameter estimation procedure.
Second, in other cases, the latent moments only provide partial information about the parameters,
but we show that this information can be naturally incorporated into the measurements framework of \citet{liang09measurements}.
This provides a natural way to leverage the method of moments when providing an
exact consistent estimator is difficult.
Finally, for factorial models, we introduce a new \emph{unshuffling
factorization} procedure to recover parameter estimates.

>>>>

% Method of moments
Recently, unsupervised learning techniques based on the method of moments and
spectral decomposition have offered a refreshing and promising perspective on
this learning problem \citep{hsu09spectral,anandkumar11tree,anandkumar12moments,anandkumar12lda,hsu12identifiability,balle11transducer,balle12automata}.
These methods exploit the linear algebraic properties of the model and
factorize the moments into parameters, providing strong theoretical guarantees.
However, these methods are not as universally applicable as EM, and have yet to
be developed for log-linear models.

\Fig{figures/schema}{0.3}{schema}{Schema of our approach.}

@@

This paper contributes three ideas:
First, in certain cases, these latent moments are exactly the sufficient statistics of
the log-linear model, in which we case use convex optimization directly to obtain the parameters,
resulting in a consistent parameter estimation procedure (\refsec{threeViewMixtureModel}).
Second, in other cases, the latent moments only provide partial information about the parameters,
but we show that this information can be naturally incorporated
using the measurements framework of \citet{liang09measurements} (\refsec{generalModels}).
This provides a natural way to leverage the method of moments when constructing
a consistent estimator is difficult.
Finally, for factorial models, we introduce a new \emph{unshuffling
factorization} procedure to recover parameter estimates (\refsec{factorialModels}).

### 2013-05-31 doc/nips2013/measurements.tex
made a pass up through section 4 (measurements)

Our learning algorithm for three-view mixture models relies on the fact that the hidden moments we can recover
are exactly the sufficient statistics of the exponential family model.
Of course, these conditions do not hold for general log-linear models.
Rather than falling back on local optimization of the likelihood in these cases, we would still like
to use the \emph{partial information} offered by the hidden moments provide we
\emph{can} recover in some way.  We will show how to incorporate these moments
into a likelihood-based objective.

Partial information is available when the following holds:
\begin{property}
  There exists some latent variable $h_{j_0} \in [\nh]$ (the bottleneck)
  and three conditionally independent observed variables $x_{j_1},x_{j_2},x_{j_3}$ (views) such that
  each conditional mean matrix $\E[\phix(x_{j}) \mid h_{j_0}] \in \R^{\nphix \times \nh}$ ($j \in \{ j_1, j_2, j_3 \}$) has full column rank $\nh$,
  for some transformation $\phix$.
\end{property}

% Examples
\reffig{generalModels} shows examples of log-linear models.
For simplicitly, assume each hidden variable $h_j \in [\nh]$ and each observed
variable $x_j \in \R^\ell$ are indicator vectors encoding one of $\ell$ possible values.

\Fig{figures/generalModels}{0.3}{generalModels}{
Three examples of undirected log-linear models.  In each of these cases,
we can recover hidden moments, which are themselves not sufficient to identify the parameters,
but provide partial information which leads to a more well-behaved likelihood function.
}

\paragraph{(a) HMM}
Following the reduction in \citet{anandkumar12moments},
take $h_2$ as the bottleneck and $x_1, x_2, x_3$ as the three views.
Define $\pi = (\BP[h_1]) \in \R^{\nh}$,
transition matrix $T = (\BP[h_3 \mid h_2]) \in \R^{\nh \times \nh}$
and observation matrix $O = (\E[x_2 \mid h_2]) \in \R^{\nphix \times \nh}$.
$\TensorFactorize$ recovers the marginal probability of the bottleneck
$T \pi$ and conditional means of the three views
$O T^\top \diag(T \pi)^{-1}$,
$O$,
and $O T$.
Note that we can recover $T = O^\dagger(OT)$ and $\pi = T^{-1} (T\pi)$,
from which we can compute the hidden moments/sufficient statistics $\mu = \E[\phi(x,h)]$.

However, this model only uses the first three observations,
which doesn't use all the information available.
Let us consider a new reduction:
Let $h_j$ be the bottleneck where $j$ is chosen uniformly at random over $\{2,
\dots, T-1\}$, where $L$ is the number of time steps; let $x_{j-1},x_j,x_{j+1}$ be the three views.\footnote{
When we actually estimate the observed moments, we simply average over the choice of $j$.}
Now, note that we can recover $O = \inv{L-2} \sum_{j=2}^{L-1} \E[x_j \mid h_j]$ from the second view,
but using the data more fully.  The first and third views are unfortunately now
quite complex due to marginalization over the rest of the variables.

\paragraph{(b) Tall mixture model}
Let $h_0$ be the bottleneck and $x_1,x_2,x_3$ be the three
views.  The conditional means that $\TensorFactorize$ reveals are $\E[x_j \mid h_0]$, which are not even
expectations over features in the original model, let alone sufficient
statistics.\footnote{Note that if the features were arbitrary, then the
parameters would not be identifiable, but because log-linear models permit
parameter sharing, then identifiability is still possible and therefore
parameter estimation in this model is still meaningful.}

\paragraph{(c) Grid model}
As in the HMM example, choose a random $h_{ij}$ as the bottleneck,
$x_{ij}^a,x_{ij}^b$ as two of the views, and any other observed variable as the third view.
$\TensorFactorize$ would reveal conditional means $\E[x_{ij}^a \mid h_{ij}]$ and $\E[x_{ij}^b \mid h_{ij}]$,
and a third quantity that involves the hidden-to-hidden features and the observation features.
Unfortunately, this quantity involves marginalizing out the rest of the graph,
and therefore impossible to extract something useful like $\BP[h_{ij}, h_{i(j+1)}]$
in closed form.
Note that the analog of this was still possible in the first HMM reduction that used only the first three observations,
but not in the second one.

In each of these three models, we could extract marginals $\pi = \BP[h_j] \in \R^{\nh}$
and some conditional mean matrix $B = \E[x_{j'} \mid h_j] \in \R^{\nphix \times \nh}$,
but that it was insufficient to identify the parameters.  How can we still use this partial information $(\pi,B)$?
To this end, we adopt the measurements framework of \citet{liang09measurements},
which allows us to incorporate moment constraints into a likelihood-based objective in a coherent way.
The basic idea is to define two families of distributions: $\sQ$ which are
consistent with all information and $\sP$ the exponential family.  The
framework essentially tries to choose $q \in \sQ$ and $p \in \sP$ to minimize
$\KL{q}{p}$.

Specifically,
let $\bx = (x^{(1)}, \dots, x^{(n)})$ be the vector of observed variables
$\bh = (h^{(1)}, \dots, h^{(n)})$ be the vector of latent variables.
Our model family $\sP$ is defined by product distributions
$p_\theta(\bx,\bh) = \prod_{i=1}^n p_\theta(x^{(i)}, h^{(i)})$ as $\theta \in \R^d$.
Suppose our observations are $(\tilde x^{(1)}, \dots, \tilde x^{(n)})$
and we have recovered $(\pi,B)$, where $\pi$ is the marginal of the bottleneck $h_j$ and $B$ is the conditional mean of $x_{j'}$.

We introduce the following constraints on each $q \in \sQ$:\footnote{We use a hard constraint
for expository clarity, but in practice, we use a quadratic penalty.  See
\citet{liang09measurements} for details.}
\begin{align}
  q(x^{(i)} = \tilde x^{(i)}) &= 1, \quad i \in [n] \label{eqn:exConstraint} \\
  \inv{n} \sum_{i \in [n]} q(x_{j'}^{(i)} = v, h_j^{(i)} = u) &= \pi_u B_{v u}, \quad u \in [\nh], v \in [\nphix]. \label{eqn:momentConstraint}
\end{align}
The first constraint $\sQ$ encodes our observed examples
and the second puts additional constraints on the latent variables.
Our final objective function is as follows:
\begin{align}
  \min_{q \in \sQ} \min_{\theta \in \R^d} \KL{q(\bx, \bh)}{p_\theta(\bx, \bh)}. \label{eqn:minKL}
\end{align}
\paragraph{Intuitions.}
With only the first constraint \refeqn{exConstraint}, the objective
\refeqn{minKL} is equivalent to our original maximum (empirical) marginal
likelihood problem, which is non-convex.  If the second constraint
\refeqn{momentConstraint} are exactly the features/sufficient statistics of the
model (as in the three-view mixture model from \refsec{threeViewMixtureModel}),
then \refeqn{minKL} is always convex and
converges to the (expected) maximum likelihood objective function as the number of examples $n \to\infty$.

>>>>

Our learning algorithm for three-view mixture models relies on the fact that
the sufficient statistics of the exponential family
can be expressed as a function of the latent moments.
Of course, these conditions do not hold for general log-linear models.
Rather than falling back on plain local optimization of the likelihood in these
cases, we would still like
to somehow use the \emph{partial information} offered by the latent moments.
%We will show how to incorporate these moments into a likelihood-based
%objective.
Partial information is available when the following holds:
\begin{property}
  There exists some latent variable $h_{j_0} \in [\nh]$ (the bottleneck)
  and three conditionally independent observed variables $x_{j_1},x_{j_2},x_{j_3}$ (views) such that
  each conditional mean matrix $\E[\phix(x_{j}) \mid h_{j_0}] \in \R^{\nphix \times \nh}$ ($j \in \{ j_1, j_2, j_3 \}$) has full column rank $\nh$
  for some transformation $\phix$.
\end{property}

% Examples
\reffig{generalModels} shows examples of log-linear models.
For notational simplicitly, assume each latent variable $h_j \in [\nh]$ is an integer
and each observed variable $x_j \in \R^\ell$ is an indicator vector encoding
one of $\ell$
possible values.

\Fig{figures/generalModels}{0.3}{generalModels}{
Three examples of undirected log-linear models.  In each of these cases,
we can recover latent moments, which are themselves not sufficient to identify the parameters,
but provide partial information which leads to a more well-behaved likelihood function.
}

\paragraph{(a) HMM}
Following the reduction in \citet{anandkumar12moments},
we could take $h_2$ as the bottleneck and $x_1, x_2, x_3$ as the three views.
Define $\pi = (\BP[h_1]) \in \R^{\nh}$,
transition matrix $T = (\BP[h_3 \mid h_2]) \in \R^{\nh \times \nh}$
and observation matrix $O = (\E[x_2 \mid h_2]) \in \R^{\nphix \times \nh}$.
$\TensorFactorize$ recovers the marginal probability of the bottleneck
$T \pi$ and conditional means of the three views
$O T^\top \diag(T \pi)^{-1}$,
$O$,
and $O T$.
Note that we can recover $T = O^\dagger(OT)$ and $\pi = T^{-1} (T\pi)$,
from which we can compute the latent moments/sufficient statistics $\mu = \E[\phi(x,h)]$.

However, this model only uses the first three observations,
which doesn't use all the information available.
Let us consider a new reduction:
Let $h_j$ be the bottleneck where $j$ is a \emph{random} variable
taking on values uniformly from $\{2,
\dots, L-1\}$, where $L$ is the number of time steps.
Let $x_{j-1},x_j,x_{j+1}$ be the three views.\footnote{
When we actually estimate the observed moments, we simply average over the choice of $j$.}
Now, note that we can recover $O = \inv{L-2} \sum_{j=2}^{L-1} \E[x_j \mid h_j]$ from the second view,
but using the data more fully.  The conditional means of the first and third
views are now a complex function of the parameters.
%due to marginalization over the rest of the variables,
However, the point is that we can get most of our utility out of the partial
information offered by $O$ and ignore information which is hard to ingest.

\paragraph{(b) Grid model}
Analogous to the second HMM reduction, let $h_{ij}$ be the bottleneck,
where $(i,j)$ is chosen randomly.
Let $x_{ij}^a,x_{ij}^b$ be two of the views, and any other observed variable as the third view.
$\TensorFactorize$ would reveal conditional means $\E[x_{ij}^a \mid h_{ij}]$
and $\E[x_{ij}^b \mid h_{ij}]$,
and a third quantity that involves the latent-to-latent features and the observation features.
This quantity involves marginalizing out the rest of the graph,
and therefore difficult to use.
%to extract something useful like $\BP[h_{ij}, h_{i(j+1)}]$
%in closed form.
%Note that the analog of this was still possible in the first HMM reduction that
%used only the first three observations, but not in the second one.

\paragraph{(c) Tall mixture model}
Let $h_0$ be the bottleneck and $x_1,x_2,x_3$ be the three
views.
$\TensorFactorize$ reveals the conditional means $\E[x_j \mid h_0]$, which are
not even expectations over features in the original model, let alone sufficient
statistics.\footnote{Note that
the model might not be identifiable in general \cite{anandkumar11tree}.
However, the model is identifiable with adequate parameter sharing,
and therefore
parameter estimation in this model is still meaningful.}

In each of these three models, we could extract marginals $\pi = \BP[h_j] \in \R^{\nh}$
and some conditional mean matrix $B = \E[x_{j'} \mid h_j] \in \R^{\nphix \times \nh}$,
but these are insufficient to identify the parameters.
How can we still use this partial information $(\pi,B)$?
To this end, we adopt the measurements framework of \citet{liang09measurements},
which allows us to incorporate moment constraints into a likelihood-based objective in a coherent way.
The basic idea is to define two families of distributions:
$\sQ$, which are consistent with all information; and
$\sP$, the exponential family.  The
framework tries to choose $q \in \sQ$ and $p \in \sP$ to minimize
$\KL{q}{p}$.

Specifically,
let $\bx = (x^{(1)}, \dots, x^{(n)})$ be the vector of observed variables
and let
$\bh = (h^{(1)}, \dots, h^{(n)})$ be the vector of latent variables.
Our model family $\sP$ is defined by product distributions
$p_\theta(\bx,\bh) = \prod_{i=1}^n p_\theta(x^{(i)}, h^{(i)})$ as $\theta \in \R^d$.
Suppose our observations are $(\tilde x^{(1)}, \dots, \tilde x^{(n)})$
and we have recovered $(\pi,B)$, where $\pi$ is the marginal of the bottleneck $h_j$ and $B$ is the conditional mean of $x_{j'}$.

Define $\sQ$ to be $q$ satisfying the following constraints:\footnote{We use a hard constraint
for expository clarity, but in practice, we use a quadratic penalty.  See
\citet{liang09measurements} for details.}
\begin{align}
  q(x^{(i)} = \tilde x^{(i)}) &= 1, \quad i \in [n] \label{eqn:exConstraint} \\
  \inv{n} \sum_{i \in [n]} q(x_{j'}^{(i)} = v, h_j^{(i)} = u) &= \pi_u B_{v u}, \quad u \in [\nh], v \in [\nphix]. \label{eqn:momentConstraint}
\end{align}
The first constraint $\sQ$ encodes our observed examples
and the second puts additional constraints on the latent variables.
Our final objective function is as follows:
\begin{align}
  \min_{q \in \sQ} \min_{\theta \in \R^d} \KL{q(\bx, \bh)}{p_\theta(\bx, \bh)}. \label{eqn:minKL}
\end{align}
\paragraph{Intuitions.}
With only the first constraint \refeqn{exConstraint}, the objective
\refeqn{minKL} is equivalent to our original maximum (empirical) marginal
likelihood problem, which is non-convex.  If the second constraint
\refeqn{momentConstraint} are exactly the features/sufficient statistics of the
model (as in the three-view mixture model from \refsec{threeViewMixtureModel}),
then \refeqn{minKL} is always convex and
converges to the (expected) maximum likelihood objective function as the number of examples $n \to\infty$.

### 2013-05-31 doc/nips2013/mixture.tex
First cut at the experiments section
made a pass up through section 4 (measurements)

In this section, we consider the simplest model to illustrate the form of our method.
Consider the undirected mixture model model in \reffig{simpleModels}(a).
We have one discrete latent variable $h \in [\nh]$ and
three observed variables
$x = (x_1, x_2, x_3) \in \sX_1 \times \sX_2 \times \sX_3 = \sX$
which are conditionally independent given $h$.
Formally, the model is as follows:\footnote{We assume for simplicity that the $x_j$'s are identically distributed conditioned on $h$,
but it is possible to use different distributions (see \citet{anandkumar12moments}).}
\begin{align}
  \label{eqn:mixtureModel}
  p_\theta(x, h) \eqdef \exp \{ \phi(x, h)^\top\theta - A(\theta) \}, \quad\text{where} \, \phi(x, h) \eqdef \phiprior(h) + \sum_{j=1}^3 \philocal(x_j, h) \in \R^d.
\end{align}

In the following, we will detail the algorithm for producing a parameter
estimate $D$.  For simplicity of exposition, we will assume we have an infinite
amount of data and work with the population quantities.  Actual parameter
estimates are constructed by simply plugging in empirical quantities.

@@

Rather than directly trying to connect the parameters $\theta$ to the data, we start
bottom-up by considering what properties of the data we can measure.
Let $\phix : \sX \mapsto \R^\ell$ denote a \emph{transformation function} (the requirement will be specified later),
which will extract the properties of the observed variables $x$ relevant to $h$.
Define the \emph{observed moments} (analogous to those used in \cite{anandkumar12moments}):
\begin{align}
  M_2 \eqdef \E[\phix(x_1) \otimes \phix(x_2)] \in \R^{\nphix \times \nphix}, \quad\quad
  M_3 \eqdef \E[\phix(x_1) \otimes \phix(x_2) \otimes \phix(x_3)] \in \R^{\nphix \times \nphix \times \nphix}.
\end{align}
Importantly, these quantities are easily estimated from data by taking empirical averages.

\subsection{Step 2: Recover hidden moments (factorization)}

By exploiting the conditional independence structure of the model, the observed
moments $M_2,M_3$ (which only reference the observed variables) admit a
factorization in terms of the \emph{hidden moments}
(which provides a window into the hidden variables).

Define
$\pi \eqdef (\BP[h=1], \dots, \BP[h=\nh])$ to be the marginal probabilities over the latent variable $h$,
and define the matrix $B \eqdef (\E[\phix(x_j) \mid h])_{h \in [\nh]} \in \R^{\ell \times k}$,
where each column is the expected transformated data conditioned on $h$.

Now, we can express the observed moments $(M_2,M_3)$ in terms of the hidden moments $(\pi,B)$:
\begin{align}
  M_2 = \sum_h \pi_h B_h \otimes B_h \in \R^{\nphix \times \nphix}, \quad\quad
  M_3 = \sum_h \pi_h B_h \otimes B_h \otimes B_h \in \R^{\nphix \times \nphix \times \nphix}.
\end{align}
\begin{assumption}
  The hidden moments matrix $B \in \R^{\nphix \times \nh}$ has full column rank $\nh$.
\end{assumption}

\citep{anandkumar13tensor} presented a robust tensor power method, which we call $\TensorFactorize$,
which converts consistent estimates of $(M_2,M_3)$ into consistent estimates of $(\pi, B)$ (up to permutation of the columns).
In brief, algorithm first computes a whitening matrix $W \in \R^{\nphix \times \nh}$ such that $W^\top M_2 W = I$,
and uses it to whiten $M_3$.  Then a robust tensor power method is applied to extract the eigenvectors of the whitened $M_3$,
which correspond to columns of $B$.
%\begin{lemma}[tensor factorization \citep{anandkumar13tensor}]
%Let $B \in \R^{\nphix \times \nh}$ be a rank $k$ matrix.
%Let $M_2 = \sum_h \pi_h B_h \otimes B_h \in \R^{\nphix \times \nphix}$ be a rank $k$ symmetric matrix.
%and $M_3 = \sum_h \pi_h B_h \otimes B_h \otimes B_h \in \R^{\nphix \times \nphix \times \nphix}$ be a rank $k$ symmetric tensor.
%There exists an polynomial time algorithm $\TensorFactorize$ that takes consistent estimates of $M_2$ and $M_3$
%and returns consistent estimates of $(\pi, B)$ (up to permutation of the columns).
%\end{lemma}

We need to specify some conditions on the transformation function $\phix$
to ensure that enough information is retained to pinpoint the parameters later.
The following condition characterizes $\phix$:
\begin{property}
  \label{property:transformation}
  The transformation $\phix$ (chosen by us) is such that $C \eqdef
  (\E[\philocal(x_j, h) \mid h])_{h \in [\nh]} \in \R^{d \times h}$ is a
  function of $B$.
\end{property}
Here, $C$ provide a view on the local features $\phi_L(x_j,h)$ which is more
directly related to the model parameters \refeqn{mixtureModel}, whereas $B$ is
what we have recovered so far.  Note that we couldn't recover $C$ directly because it depends on $h$.

However, it is not difficult to construct $\phix$ such that Property~\ref{property:transformation} holds.
Since $h$ is discrete, we can always take $\phix(x) = \flatten((\philocal(x, h))_{h \in [\nh]}) \in \R^{dk}$,
which contains $k$ blocks, one for each latent state.  We can recover a column $C_h = \E[\philocal(x_j, h) \mid h]$
by taking the block of column $B_h = \E[\flatten(\philocal(x_j, h'))_{h' \in [\nh]} \mid h]$
corresponding to $h = h'$.
Often, the local features are simply the tensor product of some function $g(x)$ and an indicator on $h$:
$\philocal(x,h) = \flatten(g(x) \otimes e_h)$.
In this case, it suffices to take $\phix(x) = g(x)$.

Finally, having retrieved $C_h$, it is easy to compute the \emph{hidden moments} $\mu$:
\begin{align}
  \mu \eqdef \E[\phi(x,h)] = \E[\phiprior(h)] + \sum_{j=1}^3 \E[\philocal(x_j,h)] = \sum_{h \in [\nh]} \pi_h (\phi_p(h) + 3 C_h).
\end{align}
In \cite{anandkumar12moments}, these hidden moments would be exactly the parameters (multinomial probabilities), but
for log-linear models, we still need to do more work to extract the parameters $\theta$.

@@

For the mixture model, the hidden moments $\E[\phi(x,h)]$ we recover are actually
are the sufficient statistics of the log-linear model.
To elaborate, let us suppose for the moment that we have infinite examples of $(x,h)$.
Let us consider the undirected mixture model.
Then we would seek to maximize the joint log-likelihood:
\begin{align}
L(\theta) \eqdef \E[\log p_\theta(x, h \mid \theta)] = {\underbrace{\E[\phi(x,h)]}_{\mu}}^\top\theta - A(\theta).
\end{align}
Note that the optimization problem here only depends on the data via $\mu$, which is usually estimated from examples of $(x,h)$.
Since $h$ is latent in our latent-variable setting, we must recover $\mu$ from latent observations.
By the moment-generating properties of the log-partition function,
the optimality condition is that $\mu = \E_\theta[\phi(x, h)]$.

We thus obtain a consistent estimator for $\theta$ assuming the exponential family is identifiable:
\begin{assumption}
The log-linear model is identifiable from fully-observed data.
Equivalently, the log-partition function $A(\theta)$ is strongly convex.
\end{assumption}
Identifiability means that the moments $\mu$ uniquely determine $\theta^*$.
In summary, factorization yields the mean parameters of the exponential family,
and we can use convex optimization to retrieve the (canonical) parameters $\theta$.

@@

\paragraph{Perturbation analysis.}
We have so far worked with exact population values $M_2,M_3,\pi,B,\mu,\theta^*$.
With finite samples, we must work the corresponding estimated quantities $\hat M_2,\hat M_3,\hat\pi,\hat B,\hat\mu,\hat\theta$.
How the error in the moments $(\hat M_2,\hat M_3)$ affects error in $(\hat\pi,\hat B)$ is well studied
(see \citet{anandkumar12moments,anandkumar13tensor}).
The hidden moments $\hat\mu$ are simply a quadratic combination of $(\hat\pi,\hat B)$,
and $\hat\theta = \nabla^{-1} A(\hat\mu)$, which is Lipschitz.
Therefore, we have a consistent estimator with polynomial sample complexity.

>>>>

In this section, we consider the simplest model to illustrate the form of our
method.  Consider the undirected mixture model model in \reffig{simpleModels}(b).
We have one discrete latent variable $h \in [\nh]$ and
three observed variables
$x = (x_1, x_2, x_3) \in \sX_1 \times \sX_2 \times \sX_3 = \sX$
which are conditionally independent given $h$.
Formally, the model is as follows:\footnote{We assume for simplicity that the $x_j$'s are identically distributed conditioned on $h$,
but it is possible to use different distributions (see \citet{anandkumar12moments}).}
\begin{align}
  \label{eqn:mixtureModel}
  p_\theta(x, h) \eqdef \exp \{ \phi(x, h)^\top\theta - A(\theta) \}, \quad\text{where} \, \phi(x, h) \eqdef \phiprior(h) + \sum_{j=1}^3 \philocal(x_j, h) \in \R^d.
\end{align}

We will now describe the algorithm for producing a parameter
estimate $\hat\theta$.  For simplicity of exposition, we will assume we have an infinite
amount of data and work with population quantities.  Actual parameter
estimates are constructed by simply plugging in the empirical versions.

@@

%Rather than directly trying to connect the parameters $\theta$ to the data $D$,
Let us start bottom-up by simply computing properties of the data.
Let $\phix : \sX \mapsto \R^\ell$ denote a \emph{transformation function}
(the requirement on $t$ will be specified later),
which aims to extract properties of $x$ relevant to $h$.
Define the \emph{observed moments} (analogous to those used in \cite{anandkumar12moments}):
\begin{align}
  M_2 \eqdef \E[\phix(x_1) \otimes \phix(x_2)] \in \R^{\nphix \times \nphix}, \quad\quad
  M_3 \eqdef \E[\phix(x_1) \otimes \phix(x_2) \otimes \phix(x_3)] \in \R^{\nphix \times \nphix \times \nphix}.
\end{align}
Importantly, these quantities are easily estimated from data by taking empirical averages.

\subsection{Step 2: Recover latent moments (factorization)}

By exploiting the conditional independence structure of the model, the observed
moments $M_2,M_3$ (which only reference the observed variables) admit a
factorization in terms of the \emph{latent moments}
(which provides a window into the latent variables) \cite{anandkumar12moments}.

Define
$\pi \eqdef (\BP[h=1], \dots, \BP[h=\nh])$ to be the marginal probabilities
over $h$,
and define the matrix $B \eqdef (\E[\phix(x_j) \mid h])_{h \in [\nh]} \in \R^{\ell \times k}$,
where each column is the expected transformated data conditioned on $h$.

Now we can express the observed moments $(M_2,M_3)$ in terms of the latent moments $(\pi,B)$:
\begin{align}
  M_2 = \sum_h \pi_h B_h \otimes B_h \in \R^{\nphix \times \nphix}, \quad\quad
  M_3 = \sum_h \pi_h B_h \otimes B_h \otimes B_h \in \R^{\nphix \times \nphix \times \nphix}.
\end{align}
\begin{assumption}
  The latent moments matrix $B \in \R^{\nphix \times \nh}$ has full column rank $\nh$.
\end{assumption}

\citep{anandkumar13tensor} presented a robust tensor power method, which we call $\TensorFactorize$,
which converts consistent estimates of $(M_2,M_3)$ into consistent estimates of $(\pi, B)$ (up to permutation of the columns).
In brief, the algorithm first computes a whitening matrix $W \in \R^{\nphix \times \nh}$ such that $W^\top M_2 W = I_{k \times k}$,
and uses it to whiten $M_3$.  Then a robust tensor power method is applied to
extract the eigenvectors of the whitened $M_3$;
unwhitening yields the columns of $B$.
%\begin{lemma}[tensor factorization \citep{anandkumar13tensor}]
%Let $B \in \R^{\nphix \times \nh}$ be a rank $k$ matrix.
%Let $M_2 = \sum_h \pi_h B_h \otimes B_h \in \R^{\nphix \times \nphix}$ be a rank $k$ symmetric matrix.
%and $M_3 = \sum_h \pi_h B_h \otimes B_h \otimes B_h \in \R^{\nphix \times \nphix \times \nphix}$ be a rank $k$ symmetric tensor.
%There exists an polynomial time algorithm $\TensorFactorize$ that takes consistent estimates of $M_2$ and $M_3$
%and returns consistent estimates of $(\pi, B)$ (up to permutation of the columns).
%\end{lemma}

We need to specify conditions on the transformation function $\phix$
to ensure that enough information is retained to pinpoint the parameters later.
The following condition characterizes $\phix$:
\begin{property}
  \label{property:transformation}
  The transformation $\phix$ (chosen by us) is such that $C \eqdef
  (\E[\philocal(x_j, h) \mid h])_{h \in [\nh]} \in \R^{d \times h}$ is a
  function of $B$.
\end{property}
Here, $C$ provide a view on the local features $\philocal(x_j,h)$,
which are
closely related to the model parameters \refeqn{mixtureModel},
and $B$ is what we have recovered so far.
%Note that we couldn't recover $C$ directly because it depends on $h$.

It is easy to construct $\phix$ such that Property~\ref{property:transformation} holds.
Since $h$ is discrete,
we can always take $\phix(x) = \flatten((\philocal(x, h))_{h \in [\nh]}) \in \R^{dk}$,
which contains $k$ blocks, one for each latent state.
Then each column $C_h \in \R^d$ is simply the $h$-th block
of the column $B_h \in \R^{kd}$.
%We can recover a column $C_h = \E[\philocal(x_j, h) \mid h]$
%Each column $C_h = \E[\philocal(x_j, h) \mid h]$
%by taking the block of column $B_h = \E[\flatten(\philocal(x_j, h'))_{h' \in [\nh]} \mid h]$
%corresponding to $h = h'$.
More simply, if
the local features are simply the tensor product of some function $g(x)$ and an
indicator on $h$ ($\philocal(x,h) = \flatten(g(x) \otimes e_h)$),
then $\phix(x) = g(x)$ suffices.

@@

In \cite{anandkumar12moments}, our latent moments $(\pi, B)$
would be the parameters, but for log-linear models,
we still need to do a bit more work.
First, note that we can express the sufficient statistics $\E[\phi(x,h)]$
of the model in terms of the quantities recovered in Step 2:
\begin{align}
  \mu \eqdef \E[\phi(x,h)] = \E[\phiprior(h)] + \sum_{j=1}^3 \E[\philocal(x_j,h)] = \sum_{h \in [\nh]} \pi_h (\phiprior(h) + 3 C_h).
\end{align}
By properties of the exponential family, $\mu$ suffices to recover $\theta$.
To elaborate, if we had infinite \emph{labeled} examples $(x,h)$,
then we would seek to maximize the joint log-likelihood:
\begin{align}
L(\theta) \eqdef \E[\log p_\theta(x, h \mid \theta)] = {\underbrace{\E[\phi(x,h)]}_{\mu}}^\top\theta - A(\theta).
\end{align}
Note that the optimization problem here only depends on the data via $\mu$.
Instead of estimating $\mu$ from labeled examples,
we estimate $\mu$ from tensor factorization applied to unlabeled examples.

We thus obtain a consistent estimator for $\theta$ assuming the exponential
family is identifiable from fully-observed data (that is
the log-artition function $A(\theta)$ is strongly convex).
%\begin{assumption}
%The log-linear model is identifiable from fully-observed data.
%Equivalently, the log-partition function $A(\theta)$ is strongly convex.
%\end{assumption}
%Identifiability means that the moments $\mu$ uniquely determine $\theta^*$.
%By the moment-generating properties of the log-partition function,
%the optimality condition is that $\mu = \nabla A(\theta) = \E_\theta[\phi(x, h)]$.
In other words,
tensor factorization (Step 2) yields the mean parameters of the exponential family
($\mu = \nabla A(\theta) = \E_\theta[\phi(x, h)]$),
which can be mapped to (canonical) parameters $\theta$ via convex optimization.

@@

\paragraph{Perturbation analysis (sketch)}
We have so far worked with exact population values $M_2,M_3,\pi,B,\mu,\theta^*$.
With finite samples, we must work the corresponding estimated quantities $\hat M_2,\hat M_3,\hat\pi,\hat B,\hat\mu,\hat\theta$.
How the error in the moments $(\hat M_2,\hat M_3)$ affects error in $(\hat\pi,\hat B)$ is well studied
(see \citet{anandkumar12moments,anandkumar13tensor}).
The latent moments $\hat\mu$ are simply a quadratic combination of $(\hat\pi,\hat B)$,
and $\hat\theta = \nabla^{-1} A(\hat\mu)$, which is Lipschitz.
Therefore, we have a consistent estimator with polynomial sample complexity.
